Measuring the double-drift illusion with hand tracking: capturing illusion resets

### 2020-09-29 Manuscript.Rmd
added comparison of tracking data with Cavanagh & Tse 2019

---
title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# The functions here allow downloading the collected data from the OSF project:
source('R/data.R')
# This file has some common helper functions:
source('R/common.R')
# The functions in here are used on the data of the bounded tracking experiments:
source('R/tracking_bounded.R')
# And finally some code for the re-tracing experiment...
source('R/retrace_onePass_V4.R')
# As well as models fit to that data:
source('R/models.R')
# And finally, code to compare this data with Sirui's PSEs
source('R/time.R')
```

@@

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location up to several times the size of the gabor. At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig. 1, bottom left) or it saturates. This long duration before resetting or saturation is well beyond the integration period of other visual effects and into the range of short-term memory. Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to spontaneous resets. Alternatively, there is a simple periodicity or stochastics process underlies the changes in perception.

Here, we examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible. Earlier results for pointing to the double-drift target [@Lisi2017] show that the illusion persists for manual pointing. However, recent work has suggested that movement-related attention may lead to resets [@Nakayama2020], so that continuous real-time tracking of the gabor's position, may prevent the illusory shift in position. The ability to manually track the perceived position will also allow us to study the resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location. Hence, we first test if the illusion occurs when people continunously track the position of the double drift stimulus. Then we try to capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus while seeing it (a tracking procedure), or after having seen it (a retrace procedure). Investigating the time and location of resets in the recorded trajectories, allows us to distinguish between a spatial limit to the illusion or a temporal limit.

@@

The stimuli were phase-changing gabor patches with a gaussian mask with an sd of 0.43 _cm_ ( or ~.6 _dva_ when in the centre of the screen and viewed from a typical distance ). The whole stimulus extended 3 times the sd of the Guassian in all four cardinal directions. The stimulus used 1.5 phases of a sine wave, so that each sine phase equals ~0.58 _cm_ or ~1.5 _dva_. By applying both real motion and phases changes of the gabor, it becomes a 'double-drift stimulus' where the perceived path deviates from the real path. The gabor changed phase at different rates, called 'internal motion' here, specified in cycles per second ( _cps_ ). The real displacements of the stimulus are along the Y axis of the monitor, keeping the X coordinate constant at the middle of the screen, so that a stimulus without internal motion would appear to move forward and backward on the horizontal display. This real motion was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that stimulus would move from either extreme to the other ( a motion we call a 'pass' ) in 2, 3 or 4 _s_, corresponding to speeds of 6.75, 4.5 and 3.75 _cm/s_ ( or roughly 7.0, 4.7 or 3.5 _dva/s_ ). For simplicity, and since the distance moved is always the same, these speeds will be refered to by the duration of a single pass (2, 3 or 4 _s_).

@@

In the two tasks in experiment 1, participants were to do online tracking (using their unseen hand / stylus) while viewing a double-drift stimulus. In experiment 2, participants first watched a double-drift stimulus and then either reproduce the perceived path of the stimulus by retracing it on the tablet, or indicate the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus. More details are given below.

@@

For experiment 1, we first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the first and last half second ( leaving 1 _s_ trajectories for the first experiment, and 3 _s_ trajectories for the second experiment). We then calculated the instantaneous heading (direction, disregarding velocity or distance) of a smooth spline interpolated version of each trajectory segment (using 101 spline interpolated position samples, to get 100 heading samples). This heading measure should depend on the illusion strength. Moreoever, when spontaneous resets are either time-dependent or space-dependent, resets should occur at roughly the same points and in the same direction in a given condition for every participant. This means that spontaneous resets should result in systematic deviations of every participant's average heading over time.

For comparison, we use the same heading measure in experiment 2. However, the individual trajectories reveal more points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect direction changes. Since the initial movement is often jittery, we only take direction changes further than ~0.39 _cm_ away from the starting point. The first (if any) of the remaining direction changes, are then taken as reset points. By averaging the x and y coordinates of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling. The average line orientation is also calculated for each of those 54 combinations of partcipants and stimulus conditions.

@@

In this experiment, two tasks were performed by two separate groups of participants. Both tasks asked participants to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion. The idea was to test both if tracking the illusion continuously would preserve the illusion and if the tracked paths reveal resets.

@@

In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass (as tracking was noisy at the inversion points) for each of the 5 internal motion speeds (normalized for the external motion). The average angle of the tracked path is given in degrees deviation from straight forward. It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ). With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001). This makes it clear that continuous tracking preserves the illusion. We will use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.

@@

The average headings are close to straight lines for all conditions. Since average straight trajectories could either result from uniform tracking paths, each with no signs of resets or from paths with many differently shaped, directed and timed resets, they do not prove or disprove neither the absence nor the presence of resets. Hence, we also show full trajectories of all participants in each of the 5 internal motion conditions.

In order to detect potential resets, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus. Since both strong, "jump" resets and weak "hit-the-wall" resets (see Fig 1, bottom left) would change percieved direction back to or crossing the veridical motion path, their smooth-spline interpolated paths would contain samples similar in heading to the 0 internal motion condition. We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset. These trajectories are drawn in red (only 1 per condition/participant), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus at the reversal points, rather than resets.

In a second task, we only used the higher internal speed (±3 _cps_) and increased the duration of a single pass of the stimulus from 2 _s_ to 4 _s_, both the increase the occurence of spontaneous resets. Average headings and heading distributions for 6 participants can be seen in Figure 3.

@@

@@

In Figure 4, for each of the 6 combinations of internal speed (2, 3 or 4 _cps_) and external speed (3 or 4 _s_ stimulus pass) we show the data for one example participant, as well as the heading distribution for all participants. While the heading distributions don't show resets, for many of the participants, we can see that the re-traced paths do show resets, many of which are detected by our algorithm.

```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig4}**Figure 4. Example Trajectories and Heading in Experiment 2.** For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show one example participant's trajectories and reset points, as well as the heading distribution."}
plotExampleData()
```

@@

First, we consider the line angle responses as a measure of illusion strength. The strength of the illusion should depend on both the internal and external speed of the double drift stimulus. As can be seen in Figure 5 (left), at least the internal speed, and perhaps the external speed may predict illusion strength as measured by line angles. To test this, we perform a mixed-design ANOVA on line angle, using both internal and external speed as within-subject factors. We find main effects of both internal speed ( _F_(2,16)=12.7; _p_<.001, $\eta^2$=0.049) and external speed ( _F_(1,8)=6.89; _p_=.03; $\eta^2$=0.031), and no interaction. That is: we verify that illusion strength varied systematically with our manipulations of the stimulus.

```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5. Measures of Illusion Strength** Line angle responses and reset point direction should both measure illusion strength. **Left:** Line Angle response as a function of internal and external speed, averaged across 9 participants in each of the 6 conditions. Shaded areas denote 95% confidence intervals. **Right:** Angles of samples halfway to a reset over line angle response. Each dot corresponds to one participants average in one of the 6 conditions. Diagonal line indicates a weighted regression (weighted by one over the square of the line angle response to counter heteroscedasticity), and the shaded area shows a 95% confidence interval for the regression."}
plotValidity()
```

We also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as an alternative measure of illusion strength. If both the angle half-way to reset and line angle responses are good measures of illusion strength, they should be related. As can be seen from the data in Figure 5 (right), there might be a relation between the two variables, but there is also heteroscadasticity. We counter this by doing a weighted regression, with the weights set to one over the squared line angle ($\frac{1}{\alpha^2}$). Reported line angles predicted reset point direction (slope=0.71; _t_(49)=4.54, _p_<.001; $R^2$=.282; _F_(1,49)=20.6), verifying that the two measures - while imperfect - are likely both indicators of double-drift illusion strength. From here on, we will only use the angle of samples half-way to detected reset points as measure of illusion strength.

@@

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 6 (left), we can see that the reset points vary more along the Y-axis than the X-axis. That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff). In fact, the standard deviation of reset point X-coordinates is about 21.9% of teh standard deviation of reset point Y-coordinates. This already indicates that a spatial limit explains about 4 times more of the variance in reset points than a temporal limit.

```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig6}**Figure 6. Reset Points and Single Limits** **Left:** The average reset points for each participant and stimulus condition displayed spatially. Y-coordinates have been converted to time, in order to normalize for the two external speeds. Transparent purple lines indicate the standard deviation of coordinates. Dashed vertical red line indicates the median X-coordinate (1.6 _cm_). Dashed horizontal blue line indicates the median Y-coordinate (1.4 _s_). **Top right:** X-coordinates predicted from a single temporal limit. **Bottom right:** Y-coordinates predicted from a single spatial limit."}
plotResetPoints()
```

@@

Figure 7 shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).

```{r fig7, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig7}**Figure 7. Reset Points and Combined Limits**"}
plotSequentialModel()
```

@@

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. In an as of yet unpublished study [Sirui] obtained the point of subjective equivalence (PSE) of a double drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared. The double-drift stimulus was on a 2 or 4 DVA length trajectory, traversed in 1, 2 or 3 seconds. The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7%/s), that is: it explains the data using only a temporal bound on the illusion. Interestingly, this fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. We will compare the two data sets here.

```{r fig8, echo=FALSE, message=FALSE, fig.width=6, fig.height=8, fig.cap="\\label{fig:fig8}**Figure 8. PSE's of double-drift position and continuous tracking**"}
plotAverageTrajectories()
```

If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then they should show a similar pattern as the averaged trajectories in experiment 2. In order to be able to compare the two data sets, we will include all trajectories, including those where our simple algorithm did not detect a reset, as such trials should also be included in the PSE data. Further, we scale the trajectories on both axes, so that the Y end-point matches the duration of stimulus. Since the PSE data was collected in degrees visual angle, we convert the unit to degrees visual angles, using the estimate at the centre of the workspace. For the 4-second stimuli, we removed the last second. Figure 8 shows the individual participants' and the averaged trajectories (purple, red, orange) as well as the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) in gray.

@@

We set out to test if spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli. To measure the location of resets we first ask participants to track the perceived location of the double-drift stimulus online. While the heading of the tracking movements is sensitive to the strength of the illusion, there are no clear reset points in the tracking data. In a second experiment we ask participants to draw the perceived path of the stimulus after a single pass. In this re-tracing experiment the drawn traces reflect the strength of the illusion just as well, and in most trials there is a clear reset point, for either a jump reset or a hit-the-wall reset. The positions of these reset points do no vary much with elapsed time, but are better explained by the distance from the true position. This has implications for the way position perception relies on motion signals.

>>>>

---
title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# The functions here allow downloading the collected data from the OSF project:
source('R/data.R')
# This file has some common helper functions:
source('R/common.R')
# The functions in here are used on the data of the bounded tracking experiments:
source('R/tracking_bounded.R')
# Here we compare the tracking data with an earlier paper:
source('R/compare.R')
# And then some code for the re-tracing experiment...
source('R/retrace_onePass_V4.R')
# As well as models fit to that data:
source('R/models.R')
# And finally, code to compare this data with Sirui's PSEs
source('R/time.R')
```

@@

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location up to several times the size of the gabor. The strength of the illusion, or how much the perceived position shifts from the real position over time, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015].

At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig. 1, bottom left) or it saturates. This long duration before resetting or saturation is well beyond the integration period of other visual effects and into the range of short-term memory. Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to spontaneous resets. Alternatively, there is a simple periodicity or stochastics process underlies the changes in perception. Recent work has shown that independent of the stimulus, movement-related attention may lead to resets [@Nakayama2020]. This means that participants' responses to the stimulus may change the illusion. Here we will first test if real-time as well as delayed, manual tracking of the gabor's position, prevents or decreases the illusory shift in position.

The ability to record trajectories the perceived position will then allow us to study spontaneous resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location. In particular, if a conflict between two location signals [@Lisi2015] leads to resets this would predict a set distance between the real and perceived position where resets occur. Under this hypothesis, the timing of spontaneous resets would depend on the speed of internal motion (given constant external motion). On the other hand, if spontaeous resets are the result of a process that integrates over time, resets would occur after a certain amount of time, but not be dependent on internal motion. Investigating the time and location of resets in the recorded trajectories, allows us to distinguish between a spatial limit to the illusion or a temporal limit. This in turn will inform us on the processes underlying the perception of positions of moving objects.

@@

The stimuli were phase-changing gabor patches with a gaussian mask with an sd of 0.43 _cm_ ( or ~.6 _dva_ when in the centre of the screen and viewed from a typical distance ). The whole stimulus extended 3 times the sd of the Guassian in all four cardinal directions. The stimulus used 1.5 phases of a sine wave, so that each sine phase equals ~0.58 _cm_ or ~1.5 _dva_. By applying both real motion and phases changes to the gabor, it becomes a 'double-drift stimulus' where the perceived path deviates from the real path. The gabor changed phase at different rates, called 'internal motion' here, specified in cycles per second ( _cps_ ). The real displacements of the stimulus are along the Y axis of the monitor, keeping the X coordinate constant at the middle of the screen, so that a stimulus without internal motion would appear to move forward and backward on the horizontal display. This real motion was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that the stimulus would move from either extreme to the other ( a motion we call a 'pass' ) in 2, 3 or 4 _s_, corresponding to speeds of 6.75, 4.5 and 3.75 _cm/s_ ( or roughly 7.0, 4.7 or 3.5 _dva/s_ ). For simplicity, and since the distance moved is always the same, these speeds will be refered to by the duration of a single pass (2, 3 or 4 _s_).

@@

In the two tasks in experiment 1, participants were to do online tracking (using their unseen hand / stylus) while viewing a double-drift stimulus. In experiment 2, participants first watched a double-drift stimulus and then either reproduced the perceived path of the stimulus by retracing it on the tablet, or indicated the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus. More details are given below.

@@

For experiment 1, we first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the first and last half second ( leaving 1 _s_ trajectories for the first experiment, and 3 _s_ trajectories for the second experiment). We then calculated the instantaneous heading (direction, disregarding velocity or distance) of a smooth spline interpolated version of each trajectory segment (using 101 spline interpolated position samples, to get 100 heading samples). This heading measure should depend on the illusion strength.

For comparison, we use the same heading measure in experiment 2. However, the individual trajectories reveal more points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect left-right direction changes. Since the initial movement is often jittery, we only take direction changes further than ~0.39 _cm_ away from the starting point. The first (if any) of the remaining direction changes, are then taken as reset points. By averaging the x and y coordinates of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling. The average line orientation is also calculated for each of those 54 combinations of partcipants and stimulus conditions.

@@

In this experiment, two tasks were performed by two separate groups of participants. Both tasks asked participants to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion. We test if tracking the illusion continuously, prevents or decreases the illusion.

@@

In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass (as tracking was noisy at the inversion points) for each of the 5 internal motion speeds (normalized for the external motion). The average angle of the tracked path is given in degrees deviation from straight forward. It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ). With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001). This makes it clear that continuous tracking does not prevent the illusion from ocurring. We will use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.

@@

In a second task, we only used the higher internal speed (±3 _cps_) and increased the duration of a single pass of the stimulus from 2 _s_ to 4 _s_, both to increase the occurence of spontaneous resets. Average headings and heading distributions for 6 participants can be seen in Figure 3. Despite lower external motion speed, the illusion is as strong as in the conditions with the same internal motion in the first task.

@@

### Illusion strength

The illusion is clearly there in both tasks, but does tracking decrease the strength of the illusion compared to simply observing it? We compare illusion strength to that reported previously [@Cavanagh2019] in Figure 4. As can be seen the averages from our data here (in red) coincide almost exactly with the illusion strength predicted by the model. That is; tracking the illusion while observing it does not change the strength of the illusion.

```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig4}**Figure 4. Illusion strength.** The strength of the illusion in degrees angle deviation from the physical path is plotted over external motion speed ( _dva/s_ ). In red are the average illusions strengths for ±3 _cps_, ±1 _cps and 0 cps. The black circles and line are the data and model from Cavanagh & Tse (2019). The grey lines are the illusion strengths their model predicts for the internal motion speeds used here."}
plotStrengthComparison()
```

### Resets

In order to detect potential resets in the first task, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus. Since both strong, "jump" resets and weak "hit-the-wall" resets (see Fig 1, bottom left) would change percieved direction back to or crossing the veridical motion path, their smooth-spline interpolated paths would contain samples similar in heading to the 0 internal motion condition. We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset. These trajectories are drawn in red (only 1 per condition/participant), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus at the reversal points, rather than resets.

The second task does not give any additional information.

@@

In Figure 5, for each of the 6 combinations of internal speed (2, 3 or 4 _cps_) and external speed (3 or 4 _s_ stimulus pass) we show the data for one example participant, as well as the heading distribution for all participants. While the heading distributions don't show resets, for many of the participants, we can see that the re-traced paths do show resets, many of which are detected by our algorithm.

```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig5}**Figure 5. Example Trajectories and Heading in Experiment 2.** For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show one example participant's trajectories and reset points, as well as the heading distribution."}
plotExampleData()
```

@@

First, we consider the line angle responses as a measure of illusion strength. The strength of the illusion should depend on both the internal and external speed of the double drift stimulus. As can be seen in Figure 6 (left), at least the internal speed, and perhaps the external speed may predict illusion strength as measured by line angles. To test this, we perform a mixed-design ANOVA on line angle, using both internal and external speed as within-subject factors. We find main effects of both internal speed ( _F_(2,16)=12.7; _p_<.001, $\eta^2$=0.049) and external speed ( _F_(1,8)=6.89; _p_=.03; $\eta^2$=0.031), and no interaction. That is: we verify that illusion strength varied systematically with our manipulations of the stimulus.

```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6. Measures of Illusion Strength** Line angle responses and reset point direction should both measure illusion strength. **Left:** Line Angle response as a function of internal and external speed, averaged across 9 participants in each of the 6 conditions. Shaded areas denote 95% confidence intervals. **Right:** Angles of samples halfway to a reset over line angle response. Each dot corresponds to one participants average in one of the 6 conditions. Diagonal line indicates a weighted regression (weighted by one over the square of the line angle response to counter heteroscedasticity), and the shaded area shows a 95% confidence interval for the regression."}
plotValidity()
```

We also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as an alternative measure of illusion strength. If both the angle half-way to reset and line angle responses are good measures of illusion strength, they should be related. As can be seen from the data in Figure 6 (right), there might be a relation between the two variables, but there is also heteroscadasticity. We counter this by doing a weighted regression, with the weights set to one over the squared line angle ($\frac{1}{\alpha^2}$). Reported line angles predicted reset point direction (slope=0.71; _t_(49)=4.54, _p_<.001; $R^2$=.282; _F_(1,49)=20.6), verifying that the two measures - while imperfect - are likely both indicators of double-drift illusion strength. From here on, we will only use the angle of samples half-way to detected reset points as measure of illusion strength.

@@

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 7 (left), we can see that the reset points vary more along the Y-axis than the X-axis. That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff). In fact, the standard deviation of reset point X-coordinates is about 21.9% of teh standard deviation of reset point Y-coordinates. This already indicates that a spatial limit explains about 4 times more of the variance in reset points than a temporal limit.

```{r fig7, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig7}**Figure 7. Reset Points and Single Limits** **Left:** The average reset points for each participant and stimulus condition displayed spatially. Y-coordinates have been converted to time, in order to normalize for the two external speeds. Transparent purple lines indicate the standard deviation of coordinates. Dashed vertical red line indicates the median X-coordinate (1.6 _cm_). Dashed horizontal blue line indicates the median Y-coordinate (1.4 _s_). **Top right:** X-coordinates predicted from a single temporal limit. **Bottom right:** Y-coordinates predicted from a single spatial limit."}
plotResetPoints()
```

@@

Figure 8 shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).

```{r fig8, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig8}**Figure 8. Reset Points and Combined Limits**"}
plotSequentialModel()
```

@@

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. In an as of yet unpublished study [Sirui Liu] obtained the point of subjective equivalence (PSE) of a double drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared. The double-drift stimulus was on a 2 or 4 DVA length trajectory, traversed in 1, 2 or 3 seconds. The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7%/s), that is: it explains the data using only a temporal bound on the illusion. Interestingly, this fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. We will compare the two data sets here (see Fig 9).

```{r fig9, echo=FALSE, message=FALSE, fig.width=6, fig.height=8, fig.cap="\\label{fig:fig9}**Figure 9. PSE's of double-drift position and continuous tracking**"}
plotAverageTrajectories()
```

If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then they should show a similar pattern as the averaged trajectories in experiment 2. In order to be able to compare the two data sets, we will include all trajectories, including those where our simple algorithm did not detect a reset, as such trials should also be included in the PSE data. Further, we scale the trajectories on both axes, so that the Y end-point matches the duration of stimulus. Since the PSE data was collected in degrees visual angle, we convert the unit to degrees visual angles, using the estimate at the centre of the workspace. For the 4-second stimuli, we removed the last second. Figure 9 shows the individual participants' and the averaged trajectories (purple, red, orange) as well as the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) in gray.

@@

We first tested if tracking the position of a double-drift stimulus changes the illusory percept of it's position, and found that it doesn't. We then set out to test if spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli. To measure the location of resets we first ask participants to track the perceived location of the double-drift stimulus online. While the heading of the tracking movements is sensitive to the strength of the illusion, there are no clear reset points in the tracking data. In a second experiment we ask participants to draw the perceived path of the stimulus after a single pass. In this re-tracing experiment the drawn traces reflect the strength of the illusion just as well, and in most trials there is a clear reset point, for either a jump reset or a hit-the-wall reset. The positions of these reset points do no vary much with elapsed time, but are better explained by the distance from the true position. This has implications for the way position perception relies on motion signals.
